Fast, Robust Classification with Applications to Periodic Variable Stars

### 2012-05-10 report/ms.tex
lots of changes, getting close to done

\section{Introduction}

@@

Figure \ref{fig:mira} displays an example of a Mira light curve. From the y-axis we can see that this source has higher amplitude than the Classical Cepheid (this is typical of the Mira class) and more sinusoidal variation (also typical). Note that the fourier methods appear to have estimated an incorrect period for this source. The true period appears to be around 161 days, half of the estimate.
\subsection{Classification Methodology}
There has been a lot of recent progress in the astronomy literature towards developing highly accurate classifiers for periodic variables \cite{debosscher2007automated,richards2011machine,dubath2011random}. The standard approach works as follows. A telescope observes a source $j$ at times $t_{1},\ldots,t_{l}$, recording flux measurements of $m_{1},\ldots,m_{l}$. Typically there are measurements of uncertainty on the flux measurements $e_{1},\ldots,e_{l}$. So each source $j$ is initially characterized by an $l \times 3$ matrix $D_j=\{(t_{i},m_{i},e_{i})\}_{i=1}^{l}$. Note that $l$, the number of times the source is observed, is different for each $D_j$. Also the time sampling of the flux measurements is irregular and different for each source. Associated with each source is a classification, such as Mira, Classical Cepheid, RR Lyrae, etc.

@@

Various extensions to the standard SVM have been proposed that seek to incorporate uncertainty in the features into the optimization problem. These formulations often take the form of (or can be interpreted as) minimizing the objective function for a worst case allocation of features in some fixed set, i.e.
\begin{equation}
\label{eq:svm_minimax}
\min_{\beta,\beta_0}  \max_{z \in \mathcal{X}(\rho)} \sum_{i=1}^n (1 - y_i(\beta^Tz_i + \beta_0))_{+} + \frac{\lambda}{2}\beta^T\beta
\end{equation}
where $\rho$ is some scalar and $\mathcal{X}(\rho)$ defines some set to which $(z_1, \ldots, z_n)$ belong. See \cite{bhattacharyya2004robust,shivaswamy2006second,ben2011chance} for discussions of some of these models.

One way to represent feature uncertainty is to specify that observations must lie within hyper-rectangles of feature space. Formally we let $X$ be an $n \times p$ matrix where row $i$ represents the center of the hyper-rectangle for observation $j$. We define $\Sigma$ to be an $n \times p$ matrix where $\Sigma_{i,j}$ is half the width of the hyper-rectangle around observation $i$, on feature $j$. Following the notation of the SVM above we can express this uncertainty model as
\begin{equation}
\mathcal{X}(\rho) = \{Z : X - p\Sigma \leq Z \leq X + \rho \Sigma\}
\end{equation}
$\rho$ is a global uncertainty parameter that shrinks or expands the hyper-rectangles uniformly. This model was considered in El Ghaoui\cite{el2003robust}. It is easy to show that for this form of $\mathcal{X}(\rho)$ one can explicity solve for the inner max in equation \eqref{eq:svm_minimax} and obtain
\begin{equation}
\label{eq:svmInit}
\min_{\beta}\sum_{i=1}^n(1-y_i(\beta_0+\beta^Tx_i)+\rho\sigma_i^T|\beta|)_++\frac{\lambda}{2} ||\beta||_2^2.
\end{equation}
where $\sigma_i$ is the $i^{th}$ row of $\Sigma$. In \cite{el2003robust}, El Ghaoui drew similarities between the $\rho\sigma_i^T|\beta|$ in this expression and standard $L_1$ regularization for SVMs. For example, if $\Sigma_{i,j} = 1 \ \forall i,j$ then then expression becomes $\rho||\beta||_1$. This interpretation of $\mathcal{X}(\rho)$ in part motivates the use of $\rho$ as a tuning parameter rather than a quantity that is fixed by the inherent uncertainty in the data.

In this work we will fit model \eqref{eq:svmInit} the periodic variable star data set. This provides an opportunity to make use of known feature uncertainty in our classifier. In addition, the connections between $\rho$ and sparsity, may allow us to obtain highly interpretatable classifiers, with few non-zero feature coefficients.

@@

%% \begin{enumerate}
%% \item measurements in the middle of the light curve get used more often
%% \item sometimes the intervals do not contain the feature obtained when using the whole light curve. this is because taking shorter sections of light curves biases certain features high or low (frequency significance features tend to go down, should probably just get rid of these anyway)
%% \end{enumerate}
%% Bootstrap sampling might be a good approach, but this would destroy cadence sensitive features. So not sure what to do. This might be something interesting to discuss in the paper: \textbf{Determining feature error intervals for time series / image data is difficult.}

%% \section{Light Curves}

%% Figures \ref{fig:mira}, \ref{fig:cepheid}, \ref{fig:rrlyrae} show the light curves (i.e. time series) of three stars. Each is of a different class. The figures show the raw light curve (time on x-axis, magnitude on y-axis) and folded light curve (phase on x-axis, magnitude on y-axis). The phase is determined by computing a period (sometimes wrong!, for example figure \ref{fig:mira}) for each star, and then looking at the remainder of time divided by the period. The folded light curve often make the structure of the function much clearer.

%% \begin{figure}[h]
%%   \begin{center}
%%     \begin{includegraphics}[scale=.5]{4000.pdf}
%%       \caption{Light curve of a RR Lyrae AB variable star.\label{fig:rrlyrae}}
%%     \end{includegraphics}
%%   \end{center}
%% \end{figure}

%% \section{Initial Classification Work}

%% \subsection{Ignoring Errors}
%% I made a data set of 5508 of these light curves of class Mira, Classical Cepheid, and RR Lyrae AB. Random forest (a type of classifier) obtained a 0.55\% error rate, without using any feature error information. So the current classes may be too easy. I could generate features for some stars that belong to classes that are more similar to each other e.g. using stars from classes RR Lyrae C and RR Lyrae D (not used here) are a lot alike and would have much higher error rates.

\subsection{The Data Sets}
We use light curves from the Optical Gravitational Lensing Survey III (OGLEIII).\footnote{Light curves can be obtained here: http://ogledb.astrouw.edu.pl/~ogle/CVS/}. Since we want to focus on studying sparsity and the relationship between $\rho$ and $\lambda$, we select subsamples from four classes of periodic variable stars and turn these four classes into two separate data sets.

The CEP dataset contains two classes of Cepheids --  Classical Cepheids and Population II Cepheids.  Each class consisted of $n=200$ observations and $p=58$ features so that each observation is hyperrectangle in $\mathbb{R}^{58}$. The RR data set consists of two classes of RR --  RR Lyrae AB and RR Lyrae D, again with $n=200$ and $p=58$.

\subsubsection{Preprocessing the data with quantile ranking}
Many SVM machine implementation normalize features to ensure that margins are comparable between features e.g. \cite{dimitriadou2011package}. We normalize our data by assigning normal quantiles to the 2n values (lower and upper bounds of the interval) for each feature. This normalization preserves order relations among intervals in the sense that if one interval is completely above another then after normalization it is completely above, and if there is some overlap before normalization there will be some overlap after. Features containing all zeroes were removed from the dataset and missing values were inferred using the median for the feature.

@@

\subsubsection{Performance}
The performance of the combined SVM model was compared to those of its degenerates by computing expected losses for multiple values of $\lambda$ and $\rho$.  For a given $(\lambda,\rho)$, the expected loss was computed by performing cross-validation in 90-10 splits so that the union of testing points across splits is the entire dataset and the intersection is disjoint. See surface plots of loss versus $(\lambda,\rho)$: for the CEP dataset, $\lambda\in \lbrace 0\rbrace\cup [0.375,0.750]$ and $\rho\in\lbrace 0\rbrace\cup [0.090,0.032]$ while for the RR dataset, $\lambda\in\lbrace 0\rbrace\cup [54,284]$ and $\rho\in\lbrace 0\rbrace\cup [0.375,0.750]$. Observe that $(\lambda^*,\rho^*)\neq (0,0)$ for both datasets, which implies that the combined model outperforms both of its degenerates. For the CEP dataset, the optimal $(\lambda^*,\rho^*)=(9,0.320)$ yielded 13.25\% classification error while for the RR dataset, $(\lambda^*,\rho^*)=(284,0.48)$ yielded 21.50\% classification error. THESE NUMBERS DON'T LOOK RIGHT, MUST RERUN analyze\_feature.m.

@@

\subsubsection{Feature selection as suggested by $||\beta||_0$ values}

>>>>

\section{Introduction}
In this work we design an efficient, robust support vector machine (SVM) algorithm by combining 1) automated tuning of the regularization parameter $\lambda$ and 2) classification of interval data. The performance of this novel SVM is tested using periodic variable star data where features are believed to lie within confidence intervals. The paper begins with a motivating example -- classification of astronomical time series, continues with a discussion of interval SVMs and efficient path algorithms for these models, and concludes with an application of interval SVMs to two data sets from astronomy.

@@

Figure \ref{fig:mira} displays an example of a Mira light curve. From the y-axis we can see that this source has higher amplitude than the Classical Cepheid (this is typical of the Mira class) and more sinusoidal variation (also typical). Note that the fourier methods appear to have estimated an incorrect period for this source. The true period appears to be around 161 days, half of the estimate.
\subsection{Classification Methodology}
There has been a lot of recent progress in the astronomy literature towards developing highly accurate classifiers for the time series generated by periodic variables \cite{debosscher2007automated,richards2011machine,dubath2011random}. The standard approach works as follows. A telescope observes a source $j$ at times $t_{1},\ldots,t_{l}$, recording flux measurements (how bright the object is) of $m_{1},\ldots,m_{l}$. Typically there are measurements of uncertainty on the flux measurements $e_{1},\ldots,e_{l}$. So each source $j$ is initially characterized by an $l$ tuples $D_j=\{(t_{i},m_{i},e_{i})\}_{i=1}^{l}$. Note that $l$, the number of times the source is observed, is different for each $D_j$. Also the time sampling of the flux measurements is irregular and different for each source. Associated with each source is a classification, such as Mira, Classical Cepheid, RR Lyrae, etc.

@@

Various extensions to the standard SVM have been proposed that seek to incorporate uncertainty in the features into the optimization problem. These formulations can often be written as minimizing the objective function for a worst case allocation of features in some fixed set, i.e.
\begin{equation}
\label{eq:svm_minimax}
\min_{\beta,\beta_0}  \max_{z \in \mathcal{X}(\rho)} \sum_{i=1}^n (1 - y_i(\beta^Tz_i + \beta_0))_{+} + \frac{\lambda}{2}\beta^T\beta
\end{equation}
where $\rho$ is some scalar and $\mathcal{X}(\rho)$ defines some set to which $(z_1, \ldots, z_n)$ belong. See \cite{bhattacharyya2004robust,shivaswamy2006second,ben2011chance} for discussions of some of these models.

One way to represent feature uncertainty is to specify that observations must lie within hyper-rectangles of feature space. Formally we let $X$ be an $n \times p$ matrix where row $i$ represents the center of the hyper-rectangle for observation $j$. We define $\Sigma$ to be an $n \times p$ matrix where $\Sigma_{i,j}$ is the relative uncertainty on observation $i$, on feature $j$. Following the notation of the SVM above we can express this uncertainty model as
\begin{equation}
\mathcal{X}(\rho) = \{Z : X - p\Sigma \leq Z \leq X + \rho \Sigma\}
\end{equation}
$\rho$ is a global uncertainty parameter that shrinks or expands the hyper-rectangles uniformly. We could think of $\rho$ as a robustness parameter which larger $\rho$ leads to more robust perturbations of the input data. Or from a probabilistic stand--point, larger $\rho$ will increase the chance that a random feature vector is in $\mathcal{X}(\rho)$, thus ensuring separation with higher probability.

This model, but without the $\frac{\lambda}{2}\beta^T\beta$ regularization, was considered in El Ghaoui \cite{el2003robust}. It is easy to show that for this form of $\mathcal{X}(\rho)$ one can explicity solve for the inner max in equation \eqref{eq:svm_minimax} and obtain
\begin{equation}
\label{eq:svmInit}
\min_{\beta}\sum_{i=1}^n(1-y_i(\beta_0+\beta^Tx_i)+\rho\sigma_i^T|\beta|)_++\frac{\lambda}{2} ||\beta||_2^2.
\end{equation}
where $\sigma_i$ is the $i^{th}$ row of $\Sigma$. In \cite{el2003robust}, El Ghaoui drew similarities between the $\rho\sigma_i^T|\beta|$ in this expression and standard $L_1$ regularization for SVMs. For example, if $\Sigma_{i,j} = 1 \ \forall i,j$ then then expression becomes $\rho||\beta||_1$. This interpretation of $\mathcal{X}(\rho)$ in part motivates the use of $\rho$ as a tuning parameter rather than a quantity that is fixed by the inherent uncertainty in the data.

In this work we will fit model \eqref{eq:svmInit} to the periodic variable star data sets. This provides an opportunity to make use of known feature uncertainty in our classifier. In addition, the connections between $\rho$ and sparsity, may allow us to obtain highly interpretatable classifiers, with few non-zero feature coefficients.

@@

\subsection{Preprocessing the data with quantile ranking}
Many SVM machine implementation normalize features to ensure that margins are comparable between features e.g. \cite{dimitriadou2011package}. We normalize our data by assigning normal quantiles to the 2n values (lower and upper bounds of the interval) for each feature. This normalization preserves order relations among intervals in the sense that if one interval is completely above another then after normalization it is completely above, and if there is some overlap before normalization there will be some overlap after. Features containing all zeroes were removed from the dataset and missing values were inferred using the median for the feature.

\subsection{The Data Sets}
We use light curves from the Optical Gravitational Lensing Survey III (OGLEIII).\footnote{Light curves can be obtained here: http://ogledb.astrouw.edu.pl/~ogle/CVS/}. Since we want to focus on studying sparsity and the relationship between $\rho$ and $C$, we choose two small data sets that will be easy to analyze.
\begin{enumerate}
\item The CEP dataset contains two classes of Cepheids --  Classical Cepheids and Population II Cepheids.  Each class consisted of $n=200$ observations and $p=58$ features.
\item The RR data set consists of two classes of RR Lyrae stars -- RR Lyrae AB and RR Lyrae D, again with $n=200$ and $p=58$.
\end{enumerate}

@@

\subsubsection{Performance}
We treat $C$ and $\rho$ as tuning parameters and grid search across them to find the optimal values. We use a grid search for $C$, rather than the path algorithm, because it gives more consistent results. This may be due to a bug somewhere in our path code. In order to assess performance on the grid, we perform 10 fold cross validation, with some initial processing to determine where the grid should be placed.

Figure \ref{fig:regularizations} shows cross-validation error rates for given C (x-axis) and $\rho$ y-axis for the RR data set (left) and the CEP data set (right). The color strip on the right shows performance when $C=\infty$, i.e. no $L_2$ regularization. In both cases, using $\rho>0$ and $C < \infty$ gives better performance than only relying on one method for regularization. In the left plot (RR data set), there is some evidence of a tradeoff in regularization between $\rho$ and $C$. By this we mean, as one increases $C$ (regularizes less using the standard penalty), the optimal hyper-rectangle size increases, suggesting this is making up for less $L_2$ regularization.

One attractive feature of using hyper-rectangles to represent uncertainty is the opportunity to obtain sparse solutions. In Figures \ref{fig:RRfeatures} (RR) and \ref{fig:CEPfeatures} (CEP) we list the features and coefficients for all features with non-zero coefficients for the models with the lowest error rate. In both cases, over half the features have coefficients exactly 0. It is also true that in both cases many feature have coefficients which are close to, but not quite 0. We might be able to obtain sparser solutions by increasing our $\rho$. The tradeoff to this, is that this would mean choosing a model that did not obtain optimal misclassification performance.

The performance of the SVM model was compared  all values of $\lambda$ and a grid search on $\rho$.  For a given $(\lambda,\rho)$, the expected loss was computed by performing cross-validation in 90-10 splits so that the union of testing points across splits is the entire dataset and the intersection is disjoint. See surface plots of loss versus $(\lambda,\rho)$: for the CEP dataset, $\lambda\in \lbrace 0\rbrace\cup [0.375,0.750]$ and $\rho\in\lbrace 0\rbrace\cup [0.090,0.032]$ while for the RR dataset, $\lambda\in\lbrace 0\rbrace\cup [54,284]$ and $\rho\in\lbrace 0\rbrace\cup [0.375,0.750]$. Observe that $(\lambda^*,\rho^*)\neq (0,0)$ for both datasets, which implies that the combined model outperforms both of its degenerates. For the CEP dataset, the optimal $(\lambda^*,\rho^*)=(9,0.320)$ yielded 13.25\% classification error while for the RR dataset, $(\lambda^*,\rho^*)=(284,0.48)$ yielded 21.50\% classification error. THESE NUMBERS DON'T LOOK RIGHT, MUST RERUN analyze\_feature.m.

@@

\begin{figure}[ht]
\begin{center}
\begin{tabular}{cc}
$\begin{array}{cc}
\multicolumn{1}{l}{\mbox{(a)}} & \multicolumn{1}{l}{\mbox{(b)}} \\ \\ \\ [-.35in]
\includegraphics[width=0.45\textwidth]{RR2_svmIntC} &
\includegraphics[width=0.45\textwidth]{CEP_svmIntC}
\end{array}$
\end{tabular}
\end{center}
\caption{(a) Performance across tuning parameters for RR data set. (b) Performance across tuning parameters for CEP data set.}
\label{fig:regularizations}
\end{figure}

\begin{figure}
\include{RR2_features}
\caption{RR features with non-zero coefficients. Some coefficient are reported here as 0 due to rounding. Choosing a slightly higher $\rho$ may have caused severl of the features with very small coefficients to become exactly 0, leading to a very sparse solution.\label{fig:RRfeatures}}
\end{figure}

\begin{figure}
\include{CEP_features}
\caption{CEP features. Features with large coefficients correspond well to features that are known to be important for separating these two classes.\label{fig:CEPfeatures}}
\end{figure}

%\subsubsection{Feature selection as suggested by $||\beta||_0$ values}
